Implementing Distributed Systems Directly with I/O Automata in ioa++

### 2011-09-26 papers/paper1/abstract.tex
Incorporating Dr. Roman's revisions.

To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
We develop a component model for asynchronous and concurrent systems based on the I/O automata formal model and dynamic composition.
We describe the design and implementation of ioa++ based on that system model and show how ioa++ can be used to develop protocols and other forms of distributed systems.

@@

\begin{IEEEkeywords}
TODO; component; formatting; style; styling;
\end{IEEEkeywords}

>>>>

To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
We develop a component model for asynchronous and concurrent systems based on the I/O automata formal model and dynamic composition.
We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.

@@

### 2011-09-26 papers/paper1/component_model.tex
Incorporating Dr. Roman's revisions.

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.

@@

I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
The absence of shared state thus ensures that each automaton can be deployed independently.
The external signature of an I/O automaton constitutes its interface and the result of interacting with an I/O automaton is well-defined since every action is atomic.
Composition is guaranteed by the model itself.

\subsection{Practical Considerations\label{practical}}

@@

As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
However, it lacks certain features needed to develop real components.
This subsection introduces these features and lays the foundation for adding the dynamics discussed in Section~\ref{dynamics}.

@@

Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children.
This definition excludes a parent from having any state or actions of its own.
Since it is often convenient to associate state and actions with a parent, we allow a parent to be the composition of its children and an anonymous child that contains the state and actions attributed to the parent.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

@@

\paragraph*{Explicit Composition}
Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.

@@

These observations create some interesting opportunities for designing and analyzing concurrent software.
Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
A high degree of fan-out (an output action being bound to many input actions) creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
However, the input effects can be applied in parallel since the state of each automaton is independent.
A high degree of fan-in (an automaton has many bound input actions) also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
A scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
Compositions of automata can be statically composed to take advantage of in-lining.
An individual automaton can be decomposed into a number of child automata to increase parallelism.

@@

\paragraph*{Example}
Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton Z and three Channel automata A, B, and C representing three network connections.
The Control automaton Z consists of the Sensor automaton S, the Filter automaton F, and the Statistics automaton T.
The Control automaton C starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter F and Statistics component T via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton F passes the filtered sample to the Control automaton Z via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton T calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton Z by issuing a request via the $(Z.request!, T.request?)$ binding and then receiving a response via the $(T.response!(u), Z.response?(u))$ binding.
The Control automaton Z contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata A, B, and C.
The parameters 0, 1, and 2, are used to route messages to Channel automata A, B, and C respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in Figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}\footnote{Use the actual paper.  I think it has something to do with databases.}.

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Consequently, we add the ability to deal with a dynamic set of automata and interactions.

\paragraph*{System actions}
The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
Create, bind, unbind, and destroy are collectively called \emph{system actions}.
The \emph{create} operation allows an automaton to create a child automaton.
The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.

System actions can fail for a variety of reasons.
A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{practical} would be violated.
An unbind fails if the binding does not exist.
Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.

@@

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the set defined in Section~\ref{practical} \emph{and} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

>>>>

\subsection{I/O Automata\label{io_automaton}}

In this subsection we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected for execution.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.

@@

@@

@@

Hierarchical decomposition, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
For convenience, we extend the notion that a parent automaton is the composition of its children to include an anonymous child that contains state and actions attributed to the parent.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

@@

\paragraph*{Explicit Composition}
Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
However, when using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.

@@

\ifjournal
These observations create some interesting opportunities for designing and analyzing concurrent software.
Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
A high degree of fan-out (an output action being bound to many input actions) creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
However, the input effects can be applied in parallel since the state of each automaton is independent.
A high degree of fan-in (an automaton has many bound input actions) also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
A scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
Compositions of automata can be statically composed to take advantage of in-lining.
An individual automaton can be decomposed into a number of child automata to increase parallelism.

@@

\paragraph*{Example}
Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton Z and three Channel automata A, B, and C representing three network connections.
The Control automaton Z consists of the Sensor automaton S, the Filter automaton F, and the Statistics automaton T.
The Control automaton C starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter F and Statistics component T via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton F passes the filtered sample to the Control automaton Z via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton T calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton Z by issuing a request via the $(Z.request!, T.request?)$ binding and then receiving a response via the $(T.response!(u), Z.response?(u))$ binding.
The Control automaton Z contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata A, B, and C.
The parameters 0, 1, and 2, are used to route messages to Channel automata A, B, and C respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in Figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.
\fi

I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
The absence of shared state thus ensures that each automaton can be deployed independently.
The external signature of an I/O automaton constitutes its interface and the result of interacting with an I/O automaton is well-defined since every action is atomic.
Composition, asynchrony, and concurrency are guaranteed by the model itself.

%% \subsection{Practical Considerations\label{practical}}

%% As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
%% However, it lacks certain features needed to develop real components.
%% This subsection introduces these features and lays the foundation for adding the dynamics discussed in Section~\ref{dynamics}.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not appropriate for \emph{implementing} a dynamic system of I/O automata.
%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.

To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and set of bindings $B$.
The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed.
Dynamics are simulated via the manipulation of activation flags.
A dynamic system implies that the system $S$ be dependent on time $t$ measured by the number of actions that have been executed.
Thus, $S(t+1)$ need not be equal to $S(t)$.
A \emph{system action} is an action that \emph{can} change the system $S$.

\paragraph*{System actions}
We define four indivisible operations for managing a dynamic system of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
The \emph{create} operation allows an automaton to create a child automaton.
The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.

System actions can fail for a variety of reasons.
A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{io_automaton} would be violated.
An unbind fails if the binding does not exist.
Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.

@@

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the set defined in Section~\ref{io_automata} \emph{and} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{io_automata} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

### 2011-09-26 papers/paper1/evaluation.tex
New introduction.
Incorporating Dr. Roman's revisions.

Ioa++ permits independent actions to be executed concurrently.
To test this capability, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
We also vary the complexity of each action to gain insight into how synchronization affects concurrent execution.

>>>>

ioa++ permits independent actions to be executed concurrently.
The goal of this evaluation is (1) to show the concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.

### 2011-09-26 papers/paper1/paper.tex
Incorporating Dr. Roman's revisions.

\title{Implementing Distributed Systems Directly with I/O Automata in ioa++}
\author{
  \IEEEauthorblockN{Justin R. Wilson and Christopher D. Gill}
  \IEEEauthorblockA{Department of Computer Science and Engineering\\
    Washington University in St. Louis\\
    Campus Box 1045\\
    One Brookings Drive\\
    St. Louis, MO 63130\\
    {wilsonj,cdgill}@cse.wustl.edu}
  \and
  \IEEEauthorblockN{Guia-Catalin Roman}
  \IEEEauthorblockA{School of Engineering\\
    University of New Mexico\\
    Centennial Engineering Center, Suite 3071\\
    1 University of New Mexico, , MSC01 1140\\
    Albuquerque, NM 87131-0001\\
    gcroman@unm.edu}
}

@@

>>>>

\title{Implementing Distributed Systems Directly with I/O Automata in ioa++}
\author{
  \IEEEauthorblockN{Justin R. Wilson and Christopher D. Gill}
  \IEEEauthorblockA{Department of Computer Science and Engineering\\
    Washington University in St. Louis\\
    Campus Box 1045\\
    One Brookings Drive\\
    St. Louis, MO 63130\\
    \{wilsonj,cdgill\}@cse.wustl.edu}
  \and
  \IEEEauthorblockN{Gruia-Catalin Roman}
  \IEEEauthorblockA{School of Engineering\\
    University of New Mexico\\
    Centennial Engineering Center, Suite 3071\\
    1 University of New Mexico, MSC01 1140\\
    Albuquerque, NM 87131-0001\\
    gcroman@unm.edu}
}

@@

\begin{IEEEkeywords}
concurrency; I/O automata; framework; components
\end{IEEEkeywords}

### 2011-09-26 papers/paper1/related_work.tex
Incorporating Dr. Roman's revisions.

ioa++ can be compared directly to the the IOA language~\cite{garland2003ioa} developed by Lynch et al.
IOA is programming language for I/O automata that can be used with formal verification tools.
Recently, a compiler for IOA has been developed~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} and used it to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programming written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls which complicates modeling and interfacing with external services; especially when a procedure call blocks.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.

>>>>

To our knowledge, only the Spectrum Simulation System of Goldman~\cite{goldman1990distributed} and the IOA language developed by Lynch et al.~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
Spectrum focuses on algorithm development through simulation.
IOA focuses on formal software development including both simulation and verification.

ioa++ can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programming written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls which complicates modeling and interfacing with external services; especially when a procedure call blocks.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.
